---
title: "R Notebook"
output: html_notebook
---

I'd like to see if there are any patterns in my running pace and if so use them to make predictions about my pace in future runs. I have about five years of running data stored in Strava, so I exported this data, deleted some irrelevant, redundant, or personal columns, and will import and clean it up here.

```{r setup}

#Let's start by loading necessary packages

library(tidyverse)

# Import the dataset

activities <- read_csv("activities.csv",
                       col_types = cols(
  `Activity ID` = col_double(),
  `Activity Date` = col_character(),
  `Activity Name` = col_character(),
  `Activity Type` = col_character(),
  `Elapsed Time` = col_double(),
  Distance = col_double(),
  Commute = col_logical(),
  `Moving Time` = col_double(),
  `Max Speed` = col_double(),
  `Elevation Gain` = col_double(),
  `Elevation Loss` = col_double(),
  `Elevation Low` = col_double(),
  `Elevation High` = col_double(),
  `Max Grade` = col_double(),
  `Average Grade` = col_double()
                                )
                       )

# I'm only intersted in running for now, so I'm going to filter out any other activities (which I haven't recorded as consistently)

activities <- activities %>%
        filter(`Activity Type` == "Run")

```

```{r}
head(activities)

tail(activities)
```

So we're left with 708 runs recorded betwwn July 17, 2015 and July 19, 2020. Some of the columns that might contribute to my speed include "Activity Date" (how old I was), "Distance," and "Elevation Gain." I could also estimate how well trained I was at any particular moment (say, how much I had run in the previous twelve weeks), and my recovery status (say, how much I had run in the previous two days). 


